---
title: "pkgcheck"
output:
  md_document:
    variant: gfm

  rmarkdown::html_vignette:
    self_contained: no
---

# pkgcheck

<!-- badges: start -->
[![R build status](https://github.com/ropensci-review-tools/pkgcheck/workflows/R-CMD-check/badge.svg)](https://github.com/ropensci-review-tools/pkgcheck/actions?query=workflow%3AR-CMD-check)
[![codecov](https://codecov.io/gh/ropensci-review-tools/pkgcheck/branch/main/graph/badge.svg)](https://codecov.io/gh/ropensci-review-tools/pkgcheck)
[![Project Status: Active](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
<!-- badges: end -->

Check whether a package is ready for submission to
[rOpenSci](https://ropensci.org)'s peer review system. The primary function
collates the output of
[`goodpractice`](https://github.com/ropensci-review-tools/goodpractice), including `R CMD
check` results, a number of statistics via the [`pkgstats`
package](https://github.com/ropensci-review-tools/pkgstats), and checks for
package structure expected for rOpenSci submissions. The output of this
function immediately indicates whether or not a package is "Ready to Submit".


## Installation

The easiest way to install this package is via the [associated
`r-universe`](https://ropensci-review-tools.r-universe.dev/ui#builds). As
shown there, simply enable the universe with

```{r options, eval = FALSE}
options (repos = c (
    ropenscireviewtools = "https://ropensci-review-tools.r-universe.dev",
    CRAN = "https://cloud.r-project.org"
))
```

And then install the usual way with,

```{r install, eval = FALSE}
install.packages ("pkgcheck")
```

Alternatively, the package can be installed by first installing either the
[remotes](https://remotes.r-lib.org) or [pak](https://pak.r-lib.org/) packages
and running one of the following lines:


```{r remotes, eval = FALSE}
remotes::install_github ("ropensci-review-tools/pkgcheck")
pak::pkg_install ("ropensci-review-tools/pkgcheck")
```

The package can then loaded for use with

```{r library, eval = FALSE}
library (pkgcheck)
```


## Setup

The [`pkgstats` package](https://github.com/ropensci-review-tools/pkgstats)
also requires the system libraries [`ctags`](https://ctags.io) and [GNU
`global`](https://www.gnu.org/software/global/) to be installed. Procedures to
install these libraries on various operating systems are described in a
[`pkgstats`
vignette](https://docs.ropensci.org/pkgstats/articles/installation.html). This
package also uses the [GitHub GraphQL API](https://developer.github.com/v4)
which requires a local GitHub token to be stored with an unambiguous name
including `GITHUB`, such as `GITHUB_TOKEN` (recommended), or `GITHUB_PAT` (for
Personal Authorization Token). This can be obtained from GitHub (via your user
settings), and stored using

```{r ghtok, eval = FALSE}
Sys.setenv ("GITHUB_TOKEN" = "<my_token>")
```

This can also be set permanently by putting this line in your `~/.Renviron`
file (or creating this if it does not yet exist). Once `pkgstats` has been
successfully installed, the `pkgcheck` package can then be loaded via
a `library` call:

```{r load, echo = TRUE, message = FALSE}
library (pkgcheck)
```


## Usage

The package primarily has one function, `pkgcheck`, which accepts the single
argument, `path`, specifying the local location of a git repository to be
analysed. The following code generates a reproducible report by first
downloading a local clone of a repository called
[`srr-demo`](https://github.com/mpadge/srr-demo), which contains the skeleton
of an [`srr` (Software Review Roclets)
package](https://github.com/ropensci-review-tools/srr), generated with the
[`srr_stats_pkg_skeleton()`
function](https://docs.ropensci.org/srr/reference/srr_stats_pkg_skeleton.html):

```{r pkgcheck, echo = TRUE, message = FALSE}
mydir <- file.path (tempdir (), "srr-demo")
gert::git_clone ("https://github.com/mpadge/srr-demo", path = mydir)
devtools::document (mydir, quiet = TRUE) # Generate documentation entries in "/man" directory
x <- pkgcheck (mydir)
```

That object has default `print` and `summary` methods. The latter can be used
to simply check whether a package is ready for submission:

```{r summary, collapse = TRUE}
summary (x)
```

A package may only be submitted when the summary contains all ticks and no
cross symbols. (These symbols are colour-coded with green ticks and red crosses
when generated in a terminal; GitHub markdown only renders them in
black-and-white.) The object returned from the `pkgcheck` function is a complex
nested list with around a dozen primary components. Full information can be
obtained by simply calling the default `print` method by typing the object name
(`x`).

## The `pkgcheck` GitHub action

The `pkgcheck` package also has an associated GitHub action in [the
`pkgcheck-action`
repository](https://github.com/ropensci-review-tools/pkgcheck-action).
You can use this action to run `pkgcheck` every time you push commits to
GitHub, just like the `rcmdcheck()` checks which can be installed and run via
[the `usethis::use_github_action_check_standard()`
function](https://usethis.r-lib.org/reference/github_actions.html). `pkgcheck`
includes an analogous function,
[`use_github_action_pkgcheck()`](https://docs.ropensci.org/pkgcheck/reference/use_github_action_pkgcheck.html),
which will download the workflow file defining the action into your local
`.github/workflows` folder. See [the `pkgcheck-action`
repository](https://github.com/ropensci-review-tools/pkgcheck-action)
for more details.

You can also add a `pkgcheck` badge, just like that `rcmdcheck` badge, that
will always reflect the current state of your repository's `pkgcheck` results.
To add a badge, copy the following line to your README file, filling in details
of the GitHub organization and repository name (`<org>` and `<repo>`,
respectively):

```{markdown}
[![pkgcheck](https://github.com/<org>/<repo>/workflows/pkgcheck/badge.svg)](https://github.com/<org>/<repo>/actions?query=workflow%3Apkgcheck)
```


## What is checked?

All current checks are listed in [a separate
vignette](https://docs.ropensci.org/pkgcheck/articles/list-checks.html).

### Summary of Check Results

Calling `summary()` on the object returned by the [`pkgcheck()`
function](https://docs.ropensci.org/pkgcheck/reference/pkgcheck.html) will
generate a checklist like that shown above. This checklist will also be
automatically generated when a package is first submitted to rOpenSci, and is
used by the editors to assess whether to process a submission. Authors must
ensure prior to submission that there are no red crosses in the resultant list.
(In the unlikely circumstances that a package is unable to pass particular
checks, explanations should be given upon submission about why those checks
fail, and why review may proceed in spite of such failures.)


### Detailed Check Results

Full details of check results can be seen by `print`-ing the object returned by
the [`pkgcheck()`
function](https://docs.ropensci.org/pkgcheck/reference/pkgcheck.html) (or just
by typing the name of this object in the console.)

The package includes an additional function,
[`checks_to_markdown()`](https://docs.ropensci.org/pkgcheck/reference/checks_to_markdown.html),
with a parameter, `render`, which can be set to `TRUE` to automatically render
a HTML-formatted representation of the check results, and open it in a browser.
The formatting differs only slightly from the terminal output, mostly through
the components of [`goodpractice`](http://docs.ropensci.org/goodpractice/)
being divided into distinct headings, with explicit statements in cases where
components pass all checks (the default screen output inherits directly from
that package, and only reports components which *do not* pass all checks).

This
[`checks_to_markdown()`](https://docs.ropensci.org/pkgcheck/reference/checks_to_markdown.html)
function returns the report in markdown format, suitable for pasting directly
into a GitHub issue, or other markdown-compatible place. (The [`clipr`
package](https://github.com/mdlincoln/clipr) can be used to copy this directly
to your local clipboard with `write_clip(md)`, where `md` is the output of
`checks_to_markdown()`.)


## Caching and running `pkgcheck` in the background

Running the [`pkgcheck`
function](https://docs.ropensci.org/pkgcheck/reference/pkgcheck.html) can
be time-consuming, primarily because the
[`goodpractice`](https://docs.ropensci.org/goodpractice) component runs
both a full `R CMD check`, and calculates code coverage of all tests. To avoid
re-generating these results each time, the package saves previous reports to
a local cache directory defined in `Sys.getenv("PKGCHECK_CACHE_DIR")`.

You may manually erase the contents of this `pkgcheck` subdirectory at any time
at no risk beyond additional time required to re-generate contents. By default
checks presume packages use `git` for version control, with checks updated only
when code is updated via `git commit`. Checks for packages that do not use
`git` are updated when any files are modified.

The first time
[`pkgcheck()`](https://docs.ropensci.org/pkgcheck/reference/pkgcheck.html) is
applied to a package, the checks will be stored in the cache directory. Calling
that function a second time will then load the cached results, and so enable
checks to be returned much faster. For code which is frequently updated, such
as for packages working on the final stages prior to submission, it may still
be necessary to repeatedly call
[`pkgcheck()`](https://docs.ropensci.org/pkgcheck/reference/pkgcheck.html)
after each modification, a step which may still be inconveniently
time-consuming. To facilitate frequent re-checking, the package also has a
[`pkgcheck_bg()`
function](https://docs.ropensci.org/pkgcheck/reference/pkgcheck_bg.html) which
is effectively identical to the main [`pkgcheck()`
function](https://docs.ropensci.org/pkgcheck/reference/pkgcheck.html), except
it runs in the background, enabling you to continue coding while checks are
running.

The [`pkgcheck_bg()`
function](https://docs.ropensci.org/pkgcheck/reference/pkgcheck_bg.html)
returns a handle to the [`callr::r_bg()`
process](https://callr.r-lib.org/reference/r_bg.html) in which the checks are
running. Typing the name of the returned object will immediately indicate
whether the checks are still running, or whether they have finished. That
handle is itself an [`R6` object](http://r6.r-lib.org/) with a number of
methods, notably including
[`get_result()`](https://callr.r-lib.org/reference/get_result.html) which can
be used to access the checks once the process has finished. Alternatively, as
soon as the background process, the normal (foreground) 
[`pkgcheck()`
function](https://docs.ropensci.org/pkgcheck/reference/pkgcheck.html) may
be called to quickly re-load the cached results.

## Prior Work

[The `checklist` package](https://github.com/inbo/checklist) for "checking
packages and R code".

## Code of Conduct

Please note that this package is released with a [Contributor Code of
Conduct](https://ropensci.org/code-of-conduct/). By contributing to this
project, you agree to abide by its terms.

## Contributors







<!-- ALL-CONTRIBUTORS-LIST:START - Do not remove or modify this section -->
<!-- prettier-ignore-start -->
<!-- markdownlint-disable -->

All contributions to this project are gratefully acknowledged using the [`allcontributors` package](https://github.com/ropenscilabs/allcontributors) following the [all-contributors](https://allcontributors.org) specification. Contributions of any kind are welcome!

### Code

<table>

<tr>
<td align="center">
<a href="https://github.com/mpadge">
<img src="https://avatars.githubusercontent.com/u/6697851?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/commits?author=mpadge">mpadge</a>
</td>
<td align="center">
<a href="https://github.com/maelle">
<img src="https://avatars.githubusercontent.com/u/8360597?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/commits?author=maelle">maelle</a>
</td>
<td align="center">
<a href="https://github.com/assignUser">
<img src="https://avatars.githubusercontent.com/u/16141871?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/commits?author=assignUser">assignUser</a>
</td>
<td align="center">
<a href="https://github.com/annakrystalli">
<img src="https://avatars.githubusercontent.com/u/5583057?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/commits?author=annakrystalli">annakrystalli</a>
</td>
</tr>

</table>


### Issue Authors

<table>

<tr>
<td align="center">
<a href="https://github.com/karthik">
<img src="https://avatars.githubusercontent.com/u/138494?u=7f13170b18fb671d819b115ed5a684ea21dd785d&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+author%3Akarthik">karthik</a>
</td>
<td align="center">
<a href="https://github.com/piyalkarum">
<img src="https://avatars.githubusercontent.com/u/48254643?u=370433a2ace6a030f2551575bc08fa53664fbd8f&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+author%3Apiyalkarum">piyalkarum</a>
</td>
<td align="center">
<a href="https://github.com/noamross">
<img src="https://avatars.githubusercontent.com/u/571752?u=49b086850e1716aa25615cea39250c51e085a5d8&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+author%3Anoamross">noamross</a>
</td>
<td align="center">
<a href="https://github.com/christophsax">
<img src="https://avatars.githubusercontent.com/u/1390827?u=ce6363f6da758d1bb85987d021cacc34a81c8837&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+author%3Achristophsax">christophsax</a>
</td>
<td align="center">
<a href="https://github.com/steffilazerte">
<img src="https://avatars.githubusercontent.com/u/14676081?u=579dde6328e94bc3787c99a42f7668a71884cd13&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+author%3Asteffilazerte">steffilazerte</a>
</td>
<td align="center">
<a href="https://github.com/phuongquan">
<img src="https://avatars.githubusercontent.com/u/38658964?u=5761c2835f0a4853c9d7fb844061056a23d29564&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+author%3Aphuongquan">phuongquan</a>
</td>
<td align="center">
<a href="https://github.com/s3alfisc">
<img src="https://avatars.githubusercontent.com/u/19531450?u=26be80705a31079d973246c98bf3b26d9131e7d3&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+author%3As3alfisc">s3alfisc</a>
</td>
</tr>


<tr>
<td align="center">
<a href="https://github.com/Bisaloo">
<img src="https://avatars.githubusercontent.com/u/10783929?u=38e3754466eaa200e20f0609709467b6331cdfbe&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+author%3ABisaloo">Bisaloo</a>
</td>
<td align="center">
<a href="https://github.com/Robinlovelace">
<img src="https://avatars.githubusercontent.com/u/1825120?u=4b78d134ed1814b0677455f45d932b3b4a6ba3a5&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+author%3ARobinlovelace">Robinlovelace</a>
</td>
<td align="center">
<a href="https://github.com/schneiderpy">
<img src="https://avatars.githubusercontent.com/u/77991319?u=4242d4c5942fced6368dd5c68221e6618092cbf8&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+author%3Aschneiderpy">schneiderpy</a>
</td>
<td align="center">
<a href="https://github.com/eliocamp">
<img src="https://avatars.githubusercontent.com/u/8617595?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+author%3Aeliocamp">eliocamp</a>
</td>
<td align="center">
<a href="https://github.com/osorensen">
<img src="https://avatars.githubusercontent.com/u/21175639?u=086f58ad19c3fa56899f53a57567902e6f716074&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+author%3Aosorensen">osorensen</a>
</td>
<td align="center">
<a href="https://github.com/KlausVigo">
<img src="https://avatars.githubusercontent.com/u/3372431?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+author%3AKlausVigo">KlausVigo</a>
</td>
<td align="center">
<a href="https://github.com/sjentsch">
<img src="https://avatars.githubusercontent.com/u/37706914?u=e75071cb33e1bafdb16a60d7b713975b6722e9d9&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+author%3Asjentsch">sjentsch</a>
</td>
</tr>


<tr>
<td align="center">
<a href="https://github.com/willgearty">
<img src="https://avatars.githubusercontent.com/u/7232514?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+author%3Awillgearty">willgearty</a>
</td>
</tr>

</table>


### Issue Contributors

<table>

<tr>
<td align="center">
<a href="https://github.com/ddbortoli">
<img src="https://avatars.githubusercontent.com/u/25244497?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+commenter%3Addbortoli">ddbortoli</a>
</td>
<td align="center">
<a href="https://github.com/dgkf">
<img src="https://avatars.githubusercontent.com/u/18220321?u=bef717254e5b877159fa712e2b8ad6952c816064&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+commenter%3Adgkf">dgkf</a>
</td>
<td align="center">
<a href="https://github.com/cboettig">
<img src="https://avatars.githubusercontent.com/u/222586?u=dfbe54d3b4d538dc2a8c276bb5545fdf4684752f&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+commenter%3Acboettig">cboettig</a>
</td>
<td align="center">
<a href="https://github.com/jhollist">
<img src="https://avatars.githubusercontent.com/u/5438539?u=815aa29d708acd16180967c0ffaf81fc64a08bf4&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+commenter%3Ajhollist">jhollist</a>
</td>
<td align="center">
<a href="https://github.com/ropensci-review-bot">
<img src="https://avatars.githubusercontent.com/u/77673839?u=f26c3eb2a3b6c74a7f100d4a707116d24d723d81&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+commenter%3Aropensci-review-bot">ropensci-review-bot</a>
</td>
<td align="center">
<a href="https://github.com/santikka">
<img src="https://avatars.githubusercontent.com/u/8639149?v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+commenter%3Asantikka">santikka</a>
</td>
<td align="center">
<a href="https://github.com/bnicenboim">
<img src="https://avatars.githubusercontent.com/u/5982330?u=ec5543c6d11255fd330fc03f5880a1d7bdefadd7&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+commenter%3Abnicenboim">bnicenboim</a>
</td>
</tr>


<tr>
<td align="center">
<a href="https://github.com/b-rodrigues">
<img src="https://avatars.githubusercontent.com/u/2998834?u=e8c69a3efe34d7e82979a661a77300678791163a&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+commenter%3Ab-rodrigues">b-rodrigues</a>
</td>
<td align="center">
<a href="https://github.com/philipp-baumann">
<img src="https://avatars.githubusercontent.com/u/21625034?u=f1fc41ef9e936f3bc8c6d38a78349890654310b8&v=4" width="100px;" alt=""/>
</a><br>
<a href="https://github.com/ropensci-review-tools/pkgcheck/issues?q=is%3Aissue+commenter%3Aphilipp-baumann">philipp-baumann</a>
</td>
</tr>

</table>

<!-- markdownlint-enable -->
<!-- prettier-ignore-end -->
<!-- ALL-CONTRIBUTORS-LIST:END -->
