# pkgreport

<!-- badges: start -->
[![R build status](https://github.com/mpadge/pkgreport/workflows/R-CMD-check/badge.svg)](https://github.com/mpadge/pkgreport/actions?query=workflow%3AR-CMD-check)
[![gitlab push](https://github.com/mpadge/pkgreport/workflows/push-to-gitlab/badge.svg)](https://github.com/mpadge/pkgreport/actions?query=workflow%3Apush-to-gitlab)
[![Travis build status](https://travis-ci.org/mpadge/pkgreport.svg?branch=master)](https://travis-ci.org/mpadge/pkgreport)
[![Project Status: Concept](https://www.repostatus.org/badges/latest/concept.svg)](https://www.repostatus.org/#concept)
<!-- badges: end -->

Plumber API to report on package structure and function. Uses functionality
provided by the [`r-lib/pkgapi` package](https://github.com/r-lib/pkgapi),
which is not on CRAN and must first be installed with
```{r pkgapi, eval = FALSE}
source("https://install-github.me/r-lib/pkgapi")
```
(or `remotes::install_github()` will also work).

```{r load, echo = FALSE, message = FALSE}
devtools::load_all (".", export_all = FALSE)
```
```{r load-fakey, eval = FALSE}
library (pkgreport)
```

Also uses the [github GraphQL API](https://developer.github.com/v4) which
requires a local github token to be stored with an unambiguous name including
`GITHUB` and maybe `QL`, if alternative `GITHIB` tokens already exist. This can
be obtained from github (via your user settings), and stored using
```{r ghtok, eval = FALSE}
Sys.setenv ("GITHUB_QL" = "<my_token>")
```
This can also be set permanently by putting this line in your `~/.Renviron`
file (or creating this if it does not yet exist).

The package also works by locally caching previously analysed packages, in
a `pkgreport` subdirectory of the location determined by
```{r, rappdirs, eval = FALSE}
rappdirs::user_cache_dir()
```
You may manually erase the contents of this subdirectory at any time at no risk.


## Usage

Set up a local host at specified port, currently only configured to serve on `localhost`:

```{r serve}
port <- 8000
ps <- serve_api (port = port)
```
The returned object, `ps`, is a [`processx`](https://github.com/r-lib/processx)
object identifying a background R process hosting the local API. This process
must be manually killed once finished, as demonstrated below.

A report on the structure of a specified software repository can then be
generated with the `pr_report()` function:

```{r demo-fakey, eval = FALSE}
u <- "https://github.com/ropensci/osmdata"
res <- pr_report (u, port = port)
res
```
```{r demo, echo = FALSE}
#saveRDS (res, "res.Rds")
readRDS ("res.Rds")
```


Finally, kill the `processx` process hosting the local API:
```{r kill}
ps
chk <- ps$kill()
ps
```


