# pkgcheck

<!-- badges: start -->
[![R build status](https://github.com/ropensci-review-tools/pkgcheck/workflows/R-CMD-check/badge.svg)](https://github.com/ropensci-review-tools/pkgcheck/actions?query=workflow%3AR-CMD-check)
[![gitlab push](https://github.com/ropensci-review-tools/pkgcheck/workflows/push-to-gitlab/badge.svg)](https://github.com/ropensci-review-tools/pkgcheck/actions?query=workflow%3Apush-to-gitlab)
[![codecov](https://codecov.io/gh/ropensci-review-tools/pkgcheck/branch/main/graph/badge.svg)](https://codecov.io/gh/ropensci-review-tools/pkgcheck)
[![Project Status: Concept](https://www.repostatus.org/badges/latest/concept.svg)](https://www.repostatus.org/#concept)
<!-- badges: end -->

Check whether a package is ready for submission to
[rOpenSci](https://ropensci.org)'s peer review system. The primary function
collates the output of
[`goodpractice`](https://github.com/mangothecat/goodpractice), including `R CMD
check` results, along with a number of statistics via the
[`pkgstats` package](https://github.com/ropensci-review-tools/pkgstats), and package
structure checks expected for rOpenSci submissions. The output of this function
immediately indicates whether or not a package is "Ready to Submit".

The [`pkgstats` package](https://github.com/ropensci-review-tools/pkgstats)
also requires the system libraries [`ctags`](https://ctags.io) and [GNU
`global`](https://www.gnu.org/software/global/) to be installed. Procedures to
install these libraries on various operating systems are described in the
[`pkgstats` README](https://docs.ropensci.org/pkgstats). This package also uses
the [GitHub GraphQL API](https://developer.github.com/v4) which requires
a local GitHub token to be stored with an unambiguous name including `GITHUB`,
such as `GITHUB_TOKEN` (recommended), or `GITHUB_PAT` (for Personal
Authorization Token). This can be obtained from GitHub (via your user
settings), and stored using

```{r ghtok, eval = FALSE}
Sys.setenv("GITHUB_TOKEN" = "<my_token>")
```

This can also be set permanently by putting this line in your `~/.Renviron`
file (or creating this if it does not yet exist). Once `pkgstats` has been
successfully installed, the `pkgcheck` package can then be loaded via
a `library` call:

```{r load, echo = FALSE, message = FALSE}
devtools::load_all(".", export_all = FALSE)
```
```{r load-fakey, eval = FALSE}
library(pkgcheck)
```


## Usage

The package primarily has one function, `pkgcheck`, which accepts the single
argument, `path`, specifying the local location of a git repository to be
analysed. The following code generates a reproducible report by first
downloading a local clone of a repository called
[`srr-demo`](https://github.com/mpadge/srr-demo), which contains the skeleton
of an [`srr` (Software Review Roclets)
package](https://github.com/ropensci-review-tools/srr), generated with the
[`srr_stats_pkg_skeleton()`
function](https://docs.ropensci.org/srr/reference/srr_stats_pkg_skeleton.html):

```{r pkgcheck, echo = TRUE, message = FALSE}
mydir <- file.path (tempdir (), "srr-demo")
gert::git_clone ("https://github.com/mpadge/srr-demo", path = mydir)
x <- pkgcheck (mydir)
```

That object has default `print` and `summary` methods. The latter can be used
to simply check whether a package is ready for submission:

```{r summary, collapse = TRUE}
summary (x)
```

A package may only be submitted when the summary contains all ticks and no
cross symbols. (These symbols are colour-coded with green ticks and red crosses
when generated in a terminal; GitHub markdown only renders them in
black-and-white.) The object returned from the `pkgcheck` function is a complex
nested list with around a dozen primary components. Full information can be
obtained by simply calling the default `print` method by typing the object name
(`x`).

## What is checked?

Calling `summary()` on the object returned by the [`pkgcheck()`
function](https://docs.ropensci.org/pkgcheck/reference/pkgcheck.html) will
generate a checklist like that shown above. This checklist will also be
automatically generated when a package is first submitted to rOpenSci, and is
used by the editors to assess whether to process a submission. Authors must
ensure prior to submission that there are no red crosses in the resultant list.
(In the unlikely circumstances that a package is unable to pass particular
checks, explanations should be given upon submission about why those checks
fail, and why review may proceed in spite of such failures.)

The full list of checks which packages are expected to pass currently includes:

1. Package must use [`roxygen2`](https://roxygen2.r-lib.org) for documentation.
2. Package must have a [`contributing.md`
   file](https://devguide.ropensci.org/collaboration.html#contributing-guide).
3. Package must have a [`CITATION` file in the `inst`
   directory](https://cran.r-project.org/doc/manuals/r-release/R-exts.html#CITATION-files).
4. Package must have a [`codemeta.json`
   file](https://devguide.ropensci.org/building.html#creating-metadata-for-your-package).
5. All exported functions must include examples in their documentation.
6. Left-assign operators must be used consistently throughout all code (so
   either all `=` or all `<-`, but not a mixture of both).
7. Package `DESCRIPTION` file must have a "URL" field.
8. Package `DESCRIPTION` file must have a "BugReports" field.
9. Package name must be available (or package must already be) on CRAN.
10. Package must have continuous integration tests.
11. Package must have test coverage of at least 75%.
12. No function should have a [cyclomatic
    complexity](https://github.com/MangoTheCat/cyclocomp) of 15 or greater.
13. `R CMD check` must generate no warnings or errors.
14. All statistical standards must be documented, as confirmed by the
    [`srr::srr_pre_submit()`
    function](https://docs.ropensci.org/srr/reference/srr_stats_pre_submit.html).

Several of these are by default only shown when they fail; absence from
a resultant checklist may be interpreted to indicate successful checks.

## Caching and running `pkgcheck` in the background

Running the [`pgkcheck`
function](https://docs.ropensci.org/pkgcheck/reference/pkgcheck.html) can
be time-consuming, primarily because the
[`goodpractice`](https://github.com/mangothecat/goodpractice) component runs
both a full `R CMD check`, and calculates code coverage of all tests. To avoid
re-generating these results each time, the package saves previous reports to
a local cache, in a `pkgcheck` subdirectory of the location determined by

```{r, rappdirs, eval = FALSE}
rappdirs::user_cache_dir()
```

You may manually erase the contents of this subdirectory at any time at no
risk beyond additional time required to re-generate contents. This default
location may also be over-ridden by setting an environmental variable named
`pkgcheck_cache_dir`. By default checks presume packages use `git` for version
control, with checks updated only when code is updated via `git commit`. Checks
for packages that do not use `git` are updated when any files are modified.

The first time
[`pkgcheck()`](https://docs.ropensci.org/pkgcheck/reference/pkgcheck.html)
is applied to a package, the checks will be stored in the cache directory.
Calling that function a second time will then load the cached results, and so
enable checks to be returned much faster. For code which is frequently updated,
such as for packages working on the final stages prior to submission, it may
still be necessary to repeatedly call
[`pkgcheck()`](https://docs.ropensci.org/pkgcheck/reference/pkgcheck.html)
after each modification, a step which may still be inconveniently
time-consuming. To facilitate frequent re-checking, the package also has
a [`pkgcheck_bg()`
function](https://docs.ropensci.org/pkgcheck/reference/pkgcheck_bg.html)
which is effectively identical to the main 
[`pkgcheck()`
function](https://docs.ropensci.org/pkgcheck/reference/pkgcheck.html),
except it runs in the background, enabling you to continue coding while checks
are running.

The [`pkgcheck_bg()`
function](https://docs.ropensci.org/pkgcheck/reference/pkgcheck_bg.html)
returns a handle to the [`callr::r_bg()`
process](https://callr.r-lib.org/reference/r_bg.html) in which the checks are
running. Typing the name of the returned object will immediately indicate
whether the checks are still running, or whether they have finished. That
handle is itself an [`R6` object](http://r6.r-lib.org/) with a number of
methods, notably including
[`get_result()`](https://callr.r-lib.org/reference/get_result.html) which can
be used to access the checks once the process has finished. Alternatively, as
soon as the background process, the normal (foreground) 
[`pkgcheck()`
function](https://docs.ropensci.org/pkgcheck/reference/pkgcheck.html) may
be called to quickly re-load the cached results.
