---
title: "Extending or modifying checks"
author: 
  - "Mark Padgham"
date: "`r Sys.Date()`"
vignette: >
  %\VignetteIndexEntry{How to extend checks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set (
    collapse = TRUE,
    warning = TRUE,
    message = TRUE,
    width = 120,
    comment = "#>",
    fig.retina = 2,
    fig.path = "README-"
)
options (repos = c (
    ropenscireviewtools = "https://ropensci-review-tools.r-universe.dev",
    CRAN = "https://cloud.r-project.org"
))

if (identical(Sys.getenv("IN_PKGDOWN"), "true")) {
    # this will have no effect if you've already installed ctags; it is
    # only really necessary for rOpenSci's document server.
    pkgstats::ctags_install (bin_dir = "/usr/local/bin", sudo = FALSE)
}
```

This vignette describes how to modify or extend the existing suite of checks
implemented by `pkgcheck`. Each of the internal checks is defined in a separate
file in the `R` directory of this package with the prefix of `check_` (or
`checks_` for files which define multiple, related checks). Checks only require
two main functions, the first defining the check itself, and the second
defining `summary` and `print` methods based on the result of the first
function. The check functions must have a prefix `pkgchk_`, and the second
functions defining output methods specifying must have a prefix
`output_pkgchk_`. These two kind of function are now described in the following
two sections. 

Both of these functions must also accept a single input parameter of a
`pkgcheck` object, by convention named `checks`. This object is a list of four
main items:

1. `pkg` which summarises data extracted from
   [`pkgstats::pkgstats()`](https://docs.ropensci.org/pkgstats/reference/pkgstats.html),
   and includes essential information on the package being checked.
2. `info` which contains information used in checks, including `info$git`
   detailing git repository information, `info$pkgstats` containing a summary
   of a few statistics generated from
   [`pkgstats::pkgstats()`](https://docs.ropensci.org/pkgstats/reference/pkgstats.html),
   along with statistical comparisons against distributions from all current
   CRAN packages, an `info$network_file` specifying a local directory to a
   [`vis.js`](https://visjs.org) visualisation of the function call network of
   the package, and an `info$badges` item containing information from GitHub
   workflows and associated badges, where available.
3. `checks` which contains a list of all objects returned from all
   `pkgchk_...()` functions, which are used as input to `output_pkgchk_...()`
   functions.
4. `meta` containing a named character vector of versions of the core packages
   used in `pkgcheck`.

`pkgcheck` objects generally also include a fifth item, `goodpractice`,
containing the results of [`goodpractice`
checks](https://github.com/MangoTheCat/goodpractice). The `checks` item passed
to each `pkgchk_...()` function contains all information on the `package`,
`info`, `meta`, and (optionally) `goodpractice` items. Checks may use any of
this information, or even add additional information as demonstrated below. The
`checks$checks` list represents the output of check functions, and may not be
used in any way within `pkgchk_...()` functions.

<details>
<summary>Click here to see structure of full `pkgcheck` object</summary>
<p>


This is the output of applying `pkgcheck` to a package generated with the
[`srr` function
`srr_stats_pkg_skeleton()`](https://docs.ropensci.org/srr/reference/srr_stats_pkg_skeleton.html),
with `goodpractice = FALSE` to suppress that part of the results.

```{r pkgcheck-dummy, echo = FALSE, message = FALSE, results = "hide", eval = TRUE}
d <- srr::srr_stats_pkg_skeleton (pkg_name = "dummypkg")
roxygen2::roxygenise (d)
checks <- pkgcheck::pkgcheck (d, goodpractice = FALSE)
checks$goodpractice <- "Output of goodpractice goes here"
```
                       
```{r pkgcheck-str, echo = FALSE}
str (checks)
```

</p></details>



## 1. The check function



An example is the
check for whether a package has a citation, [defined in
`R/check_has_citation.R`](https://github.com/ropensci-review-tools/pkgcheck/blob/main/R/check-has-citation.R):

```{r, cache = FALSE, echo = FALSE}
knitr::read_chunk ("../R/check-has-citation.R")
knitr::read_chunk ("../R/check-scrap.R")
```
```{r pkgchk-citation}
```

This check is particularly simple, because a `"CITATION"` file [must have
exactly that name, and must be in the `inst`
sub-directory](https://cran.r-project.org/doc/manuals/R-exts.html#CITATION-files).
This function returns a simple logical of `TRUE` if the expected `"CITATION"`
file is present, otherwise it returns `FALSE`. This function, and all functions
beginning with the prefix `pkgchk_`, will be automatically called by the main
`pkgcheck()` function, and the value stored in `checks$checks$has_citation`.
The name of the item within the `checks$checks` list is the name of the
function with the `pkgchk_` prefix removed.

A more complicated example is the function to check whether a package contains
files which should not be there -- internally called "scrap" files. The check
function itself, [defined in
`R/check-scrap.R`](https://github.com/ropensci-review-tools/pkgcheck/blob/main/R/check-scrap.R),
checks for the presence of files matching an internally-defined list including
files used to locally cache folder thumbnails such as `".DS_Store"` or
`"Thumbs.db"`. The function returns a character vector of the names of any
"scrap" files which can be used by the `print` method to provide details of
files which should be removed. This illustrates the first general principle of
these check functions; that,

::: {.alert .alert-info}
- *Any information needed when summarising or printing the check result should
  be returned from the main check function.*
:::


A second important principle is that,

::: {.alert .alert-info}
- *Check functions should never return `NULL`, rather should always return an
  empty vector (such as `integer(0)`)*.
:::

The following section considers how these return values from check functions
are converted to `summary` and `print` output.

## 2. The output function

All `output_pkgchk_...()` functions must also accept the single input parameter
of `checks`, in which the `checks$checks` sub-list will already have been
populated by calling all `pkgchk_...()` functions described in the previous
section. The `pkgchk_has_citation()` function will create an entry of
`checks$checks$has_citation` which contains the binary flag indicating whether
or not a `"CITATION"` file is present. Similarly, the [the `pkgchk_has_scrap()`
function](https://github.com/ropensci-review-tools/pkgcheck/blob/main/R/check-scrap.R)
will create `checks$checks$has_scrap` which will contain names of any scrap
files present, and a length-zero vector otherwise.

The `output_pkgchk_has_citation()` function then looks like this:

```{r output-pkgchk-citation}
```

The first lines are common to all `output_pkgchk_...()` functions, and define
the generic return object. This object must be a list with the following three
items:

1. `check_pass` as binary flag indicating whether or not a check was passed;
2. `summary` containing text used to generate the `summary` output; and
3. `print` containing information used to generate the `print` output, itself a
   `list` of the following items:
    - A `message` to display at the start of the `print` result; and
    - An `object` to be printed, such as a vector of values, or a `data.frame`.

`summary` and `print` methods may be suppressed by assigning values of `""`.
The above example of `pkgcheck_has_citation` has `print = ""`, and so no
information from this check will appear as output of the `print` method. The
`summary` field has a value that is specified for both `TRUE` and `FALSE`
values of `check_pass`. The value is determined by the result of the main
`pkgchk_has_citation()` call, and is converted into a green tick if `TRUE`, or
a red cross if `FALSE`.

Checks for which `print` information is desired require a non-empty `print`
item, as in the [`output_pkgchk_has_scrap()`
function](https://github.com/ropensci-review-tools/pkgcheck/blob/main/R/check-scrap.R):

```{r output-pkgchk-scrap}
```

In this case, both `summary` and `print` methods are only triggered `if
(!out$check_pass)` -- so only if the check fails. The `print` method generates
the heading specified in `out$print$message`, with any vector-valued objects
stored in the corresponding `obj` list item displayed as formatted lists.
A package with "scrap" files, `"a"` and `"b"`, would thus have `out$print$obj
<- c ("a", "b")`, and when printed would look like this:

```{r scrap-out, echo = FALSE}
cli::cli_alert_danger ("Package contains the following unexpected files:")
cli::cli_ul ()
cli::cli_li (c ("a", "b"))
cli::cli_end()
```

This formatting is also translated into corresponding markdown and HTML
formatting in [the `checks_to_markdown()`
function](https://github.com/ropensci-review-tools/pkgcheck/blob/main/R/format-checks.R).

The design of these `pkgchk_` and `output_pkgchk_` functions aims to make the
package readily extensible, and we welcome discussions about developing new
checks. The primary criterion for new package-internal checks is that they must
be of very general applicability, in that they should check for a condition
that *almost* every package should or should not meet.

The package also has a mechanism to easily incorporate more specific,
locally-defined checks, as explored in the following section.

## 3. Creating new checks

(Coming soon ...)
