---
title: "Extending or modifying checks"
author: 
  - "Mark Padgham"
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: true
        toc_float: true
        number_sections: false
        theme: flatly
vignette: >
  %\VignetteIndexEntry{How to extend checks}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  warning = TRUE,
  message = TRUE,
  width = 120,
  comment = "#>",
  fig.retina = 2,
  fig.path = "README-"
)
```

This vignette describes how to modify or extend the existing suite of checks
performed by `pkgcheck`. It is primarily based around the example of the
[`pkg_has_scrap()`
function](https://github.com/ropensci-review-tools/pkgcheck/blob/d02dcb630e5f38f599b6186580001510ffc5b314/R/misc-checks.R#L347-L368)
contributed by Maëlle Salmon ([\@maelle](https://github.com/maelle)) via [this
pull request](https://github.com/ropensci-review-tools/pkgcheck/pull/27).

## 1. Create a new check

New checks can be placed anywhere within the current code, including in new
files within the `R` directory. The [`pkg_has_scrap()`
function](https://github.com/ropensci-review-tools/pkgcheck/blob/d02dcb630e5f38f599b6186580001510ffc5b314/R/misc-checks.R#L347-L368)
is relatively short, and so was appended to the [`R/misc-checks.R`
file](https://github.com/ropensci-review-tools/pkgcheck/blob/main/R/misc-checks.R).
Importantly check functions may not return `NULL`. If a `NULL`-like return
value is desired, it should be a zero-length vector of appropriate type. For
example, the [`pkg_has_scrap()`
function](https://github.com/ropensci-review-tools/pkgcheck/blob/d02dcb630e5f38f599b6186580001510ffc5b314/R/misc-checks.R#L347-L368)
returns the names of any "scrap" files included within a git repository
(implying files that should generally not be committed and included). If none
are included, the function returns `character(0)` and not `NULL`.

The values returned are used to construct messages issued by `pkgcheck` in
response to these checks, so it is generally better to return actual values,
rather than generic `TRUE`, `FALSE`, or other symbolic values. This point is
illustrated below in the demonstration of the `print()` method.


## 2. Add check function to `pkgcheck` output

A new check function can be added to the result returned by [the `pkgcheck()`
function](https://github.com/ropensci-review-tools/pkgcheck/blob/main/R/pkgcheck-fn.R)
by inserting it below the commented line, `# ----- Add new checks here ---`.
The `pkg_has_scrap()` function is included within that section to illustrate.
New checks inserted in that location will by default appear in a "Miscellaneous
checks" section of the report. Checks may be inserted in other locations by
inserting them elsewhere within the sequence specified in this [main
`pkgcheck()`
function](https://github.com/ropensci-review-tools/pkgcheck/blob/main/R/pkgcheck-fn.R).

## 3. Modify tests to include new check

Tests may be updated as soon as new items are returned by the [main
`pkgcheck()`
function](https://github.com/ropensci-review-tools/pkgcheck/blob/main/R/pkgcheck-fn.R).
This generally requires the two simple modifications of:

1. Adding the name of the new test to the [list of "items" expected in the
   result of the main `pkgcheck`
   call](https://github.com/ropensci-review-tools/pkgcheck/blob/a3507752b00b2206aee34ecbe99424d4cadb59f8/tests/testthat/test-pkgcheck.R#L19-L33).
2. Increase the [expected length of the result of
   `pkgcheck()`](https://github.com/ropensci-review-tools/pkgcheck/blob/a3507752b00b2206aee34ecbe99424d4cadb59f8/tests/testthat/test-bg.R#L29)
   by one for each new test. 

## 4. Add check result to summary output

All `pkgcheck` objects have a `summary` method which generates the screen print
indicating whether or not all expected checks have been passed, and whether or
not a package is ready to be submitted for peer review to rOpenSci. The printed
summary is first generated by a function called
[`collate_checks()`](https://github.com/ropensci-review-tools/pkgcheck/blob/main/R/collate-checks.R),
which calls a series of sub-functions, generally one for each kind of check.
New checks require new sub-functions, as can be seen by [the call to
`collate_scrap_checks
(checks)`](https://github.com/ropensci-review-tools/pkgcheck/blob/a3507752b00b2206aee34ecbe99424d4cadb59f8/R/collate-checks.R#L29).
New "collate" methods should all be inserted within this single
[`collate-checks.R`
file](https://github.com/ropensci-review-tools/pkgcheck/blob/main/R/collate-checks.R),
and should be prefixed with `collate_`.

A collate method must begin within a least two [`roxygen2` comment
lines](https://roxygen2.r-lib.org), the second of which is `#' @noRd` to
suppress translation to a documentation (`.Rd`) file. The first of these is
explained at the top of that file, and should be either 
```{r tick-or-cross}
#' @return Tick or cross
```
or,
```{r cross-only}
#' @return Cross only
```
(with optional additional explanatory text). The first option is generally used
for more important tests, and indicates that a green tick or red cross will
appear to indicate that the test is respectively either passed or failed. Some
tests only prompt a red cross when they fail, and do not appear in the summary
list when they pass. These specify return values with the second option, `#'
@return Cross only`.

The [`collate_scrap_checks()`
function](https://github.com/ropensci-review-tools/pkgcheck/blob/a3507752b00b2206aee34ecbe99424d4cadb59f8/R/collate-checks.R#L310-L323)
has "Cross only", and returns `NULL` is the test is passed. If the test is
failed, it appends a simple message to the initial symbols, `paste0 ("-",
symbol_crs ())`, where `symbol_crs()` encodes the red cross, and `symbol_tck()`
the green tick for passing checks. The message should generally be short and
simple. Full information returned from the actual check function should
generally be used only in the `print` method (described  below), and not in the
`summary` method.

Finally, the name of the sub-function should then be added to the sequence of
function calls in the main [`collate_checks()`
function](https://github.com/ropensci-review-tools/pkgcheck/blob/main/R/collate-checks.R),
in which the section of "Miscellaneous Checks" is clearly indicated.


## 5. Add new check result to print output

As described above, summary checks generally only print a short, fixed message
indicating whether or test a check has been passed. Detailed returns by the
actual check function are generally only shown via the [`print`
method](https://github.com/ropensci-review-tools/pkgcheck/blob/main/R/pkgcheck-methods.R)
associated with `pkgcheck` objects. The `print` method for any new check items
is generally implemented by constructing an associated function; the
`pkg_has_scrap()` function, for example, is called
[`print_scrap()`](https://github.com/ropensci-review-tools/pkgcheck/blob/a3507752b00b2206aee34ecbe99424d4cadb59f8/R/pkgcheck-methods.R#L183-L192).
That function demonstrates that all output uses the [`cli`
package](https://cli.r-lib.org), including in this example the ["list item"
function (`cli_li`)](https://cli.r-lib.org/reference/cli_li.html).

The `print` methods generally hand-code the different responses for passing and
failing checks. The [`pkg_has_scrap()`
function](https://github.com/ropensci-review-tools/pkgcheck/blob/d02dcb630e5f38f599b6186580001510ffc5b314/R/misc-checks.R#L347-L368)
only returns values when it fails, so only requires a `print` method on check
failures. The pass (tick) and fail (cross) symbols are translated into
corresponding `cli` symbols as follows:

`summary` symbol | `print` symbol
--- | ---
symbol_crs() | cli::cli_alert_danger()
symbol_tck() | cli::cli_alert_success()

See other functions in the main [`pgkcheck-methods.R`
file](https://github.com/ropensci-review-tools/pkgcheck/blob/main/R/pkgcheck-methods.R)
for examples.

## 6. Convert new check result to markdown output

The `summary` and `print` methods described above are used to control the
appearance of checks on local terminal screens. The `pkgcheck` package also
includes a function to convert checks to markdown format, the most important
purpose of which is to render the checks for inserting within GitHub issues.
For any new checks to appear in the reports generated by `ropensci-review-bot`,
they need to be added to the [`checks_to_markdown()`
function](https://github.com/ropensci-review-tools/pkgcheck/blob/main/R/format-checks.R).
The place to insert new functions is indicated in the `checks_to_markdown()`
function by the comment line, `# ----- Add new checks here ----`, similar to
how new checks are added to the main `pgkcheck()` function described in the
second section above.

The format of these functions is best illustrated by example, with the
conversion function for the scrap checks looking like this:

```{r scrap_checks_md, eval = FALSE}
scrap_checks_md <- function (checks) {

    if (length (checks$scrap) == 0L)
        return (NULL)

    c ("",
       ":heavy_check_mark: Package contains the following unexpected files:",
       "",
       paste0 ("- ", checks$scrap),
       "")
}
```

This function returns nothing if there are no scrap files, as specified by the
`roxygen2` comments of `#' @return cross only` in the corresponding function in
[the `collate_checks.R`
file](https://github.com/ropensci-review-tools/pkgcheck/blob/main/R/collate-checks.R)),
described in section 4, above. Packages which contain scrap files will then
produce the message specified by that function, including the name of each
scrap file as a list item.

Any new functions must ensure that they return `NULL` when the nominated
component of the `checks` list is itself `NULL` -- do not presume that
components processed within a given markdown conversion function actually
exist. Note also that the text output of these markdown-formatted checks
generally starts and ends with a blank line (`""`). These lines are important
to ensure appropriate rendering of HTML components.

Finally, note that the green ticks and red crosses need to be converted to
a different format, using the following markdown conversions:

`summary` symbol | markdown symbol
--- | ---
symbol_crs() | :heavy_multiplication_x:
symbol_tck() | :heavy_check_mark:
